% tut_optimiza.tex $ this file belongs to the Molcas repository $*/
\section{ALASKA and SLAPAF: A Molecular Structure Optimization}
\label{TUT:sec:structure}
\index{Optimization}\index{Geometry}

One of the most powerful functions of {\it ab initio} calculations is geometry 
predictions.  The minimum energy structure of a molecule for a given method and 
basis set is instructive especially when experiment is unable to determine the 
actual geometry. \molcas\ performs a geometry optimization with analytical 
gradients at the SCF or RASSCF level of calculation, and with numerical 
gradients at the CASPT2 level.

In order to perform geometry optimization an input file must contain
a loop, which includes several calls: calculation of integrals (\program{SEWARD}),
calculation of energy (\program{SCF}, \program{RASSCF}, \program{CASPT2}), calculation of gradients
(\program{ALASKA}), and calculation of the new geometry (\program{SLAPAF}). 

This is an example of such input
\begin{sourcelisting}
  &GATEWAY
   coord= file.xyz
   basis= ANO-S-MB
>> EXPORT MOLCAS_MAXITER=25
>> Do While  <<
  &SEWARD
  &SCF  
  &SLAPAF
>> EndDo << 
\end{sourcelisting}

The initial coordinates will be taken from xyz file file.xyz, and the geometry 
will be optimized at the SCF level in this case. After the wave function calculation,
calculation of gradients is required, although code \program{ALASKA} is automatically
called by \molcas. \program{SLAPAF} in this case required the calculation of an
energy minimum (no input). Other options are transition states (\keyword{TS}), minimum energy
paths (\keyword{MEP-search}), etc
The loop will be terminated if the geometry 
converges, or maximum number of iterations (MaxIter) will be reached (the
default value is 50).

There are several EMIL commands 
\ifmanual
(see sect~\ref{UG:sec:emil_commands})
\fi
, which can be 
used to control geometry optimization. For example, it is possible to execute 
some \molcas\ modules only once:
\begin{sourcelisting}
>> IF ( ITER = 1 )
* this part of the input will be executed only during the first iteration
>> ENDIF
\end{sourcelisting}

Program \program{SLAPAF} is tailored to use  analytical or numerical gradients produced
by \program{ALASKA} to relax the geometry of a molecule towards an energy
minimum (default, no input required then) or a transition state. The program is also used for
finding inter state crossings (ISC), conical intersections (CI),
to compute reaction paths, intrinsic reaction coordinate (IRC) paths, etc.
%Examples as how to use the \program{SLAPAF} code is displayed in following section~\ref{TUT:sec:structure}.

\subsection{SLAPAF - Basic and Most Common Keywords}
\begin{keywordlist}
\item[TS] Computing a transition state
\item[FindTS] Computing a transition state with a constraint
\item[MEP-search] Computing a steepest-descent minimum reaction path
\item[ITER] Number of iterations
\item[INTErnal] Definition of the internal coordinates
%\item[IRC ] Intrinsic reaction coordinate analysis of a TS
\item[]

%--
\end{keywordlist} 
