% molden.tex $ this file belongs to the Molcas repository $

\section{Writing LUSCUS/MOLDEN input}
\label{UG:sec:Molden}
\index{Tool!LUSCUS} \index{LUSCUS}

By default the \program{GUESSORB}, \program{SCF}, \program{MBPT2}, \program{RASSCF},
\program{SLAPAF}, \program{LOCALISATION}, and \program{MCLR} modules
generate input in Molden format. The \program{SCF}, \program{MBPT2}, \program{RASSCF},
and \program{LOCALISATION} modules generate input for molecular orbital
analysis, \program{SLAPAF} for geometry optimization analysis, minimum energy paths,
Saddle optimization paths and IRC TS analysis,
and the \program{MCLR} module generates input for
analysis of harmonic frequencies.

The generic name of the input file and the actual
name are different for the nodes as a reflection on the data generated
by each module. Hence, the actual names (generic name) for the Molden files in each module are
\begin{itemize}
\item GUESSORB module:
\$Project.guessorb.molden (MD\_GSS)
\item SCF module:
\$Project.scf.molden (MD\_SCF)
\item MBPT2 module:
\$Project.mp2.molden (MD\_MP2)
\item RASSCF module:
\$Project.rasscf.molden (MD\_CAS) for the state-averaged natural orbitals, and
\$Project.rasscf.x.molden (MD\_CAS.x) for the state-specific natural spin orbitals,
where x is the index of a CI root.
\item SLAPAF module:
\$Project.geo.molden (MD\_GEO) for geometry optimizations,
\$Project.mep.molden (MD\_MEP) for minimum energy paths,
\$Project.irc.molden (MD\_IRC) for IRC analysis of a TS, and
\$Project.saddle.molden (MD\_SADDLE) for Saddle method TS optimizations.
\item LOCALISATION module:
\$Project.local.molden (MD\_LOC)
\item MCLR module:
\$Project.freq.molden (MD\_FREQ)
\end{itemize}

Files in MOLDEN format can be visualized by Molden program,
or by \molcas\ GUI program \program{LUSCUS}. 

Note, the MOLDEN code cannot handle cases in which some shells, with the
same angular momentum,
are Cartesian and some are real spherical harmonics. All
shells for a give angular momentum have to be either Cartesian or
real spherical harmonics, otherwise no MOLDEN file for orbitals will be generated.

For further details with respect to Molden consult
http://www.caos.kun.nl/\~{}shaft/molden/molden.html.


For further details with respect to LUSCUS consult
https://sourceforge.net/projects/luscus/.


