%! AE_library_structure.tex $ this file belongs to the Molcas repository $*/
\subsubsection{Structure of the all electron basis set library}
\label{UG:sec:structure_of_the_ae_basis_set_libraries}
 
The start of a given basis set in the library is given by the line

\namelist{/label}

where ``label'' is the basis set label, as defined below in the input
description to
\program{SEWARD}. Then follows two lines with the
appropriate literature reference for that basis set. These
cards are read by
\program{SEWARD}
and must thus be included in the library, and may not be blank.
Next is a set of comment lines, which begin with an asterisk in column 1,
giving some details of the basis
sets. A number of lines follow, which specifies the basis set:
\begin{enumerate}
\item Charge of the atom and the highest angular momentum.  
For each angular momentum (l) then follows.
\item Number of primitives and contracted functions for angular momentum l
(must be identical to those given in the basis set label) \label{it:ae-2}.
\item Exponents of the primitive functions \label{it:ae-3}.
\item The contraction matrix (with one CGTO per column). Note that all
basis sets are given in the generally contracted format, even if they
happen to be segmented. Note that the number of CGTOs must correspond
to the data given in the label \label{it:ae-4}.
\end{enumerate}

The following is an example of an entry in a basis set library.
\begin{sourcelisting}
* This is the Huzinaga 5s,2p set contracted to 3s,2p        -- Comment   
* according to the Dunning paper.                           -- Comment   
/H.TZ2P.Dunning.5s2p.3s2p.                                  -- Label            
Exponents  : S. Huzinaga, J. Chem. Phys., 42, 1293(1965).   -- First ref line   
Coefficients: T. H. Dunning, J. Chem. Phys., 55, 716(1971). -- Second ref line  
 1.0 1                                                      -- Charge, sp       
 5 3                                                        -- 5s->3s           
 52.56 7.903 1.792 0.502 0.158                              -- s-exponents  	   
 0.025374  0.0  0.0				 	    -- contr. matrix	
 0.189684  0.0  0.0				 	    -- contr. matrix	
 0.852933  0.0  0.0				 	    -- contr. matrix	
 0.0	   1.0  0.0				 	    -- contr. matrix	
 0.0	   0.0  1.0				 	    -- contr. matrix	
 2  2                                                       -- 2p->2p           
 1.5  0.5                                                   -- p-exponents       
 1.0 0.0					            -- contr. matrix    
 0.0 1.0					            -- contr. matrix    
\end{sourcelisting}
